---
title: "Culminating Activity"
author: Avryl Joie Arranguez, Dustin Jesse Balansag, Leira Bengil, John Andrew Chua,
  Jian Bryce Machacon
date: "2026-05-03"
output:
  html_document:
    df_print: paged
---
# Initializing & inspecting the Dataset
```{r}
setwd("/home/matchacone/Downloads/Data-Analytics-Culminating-Activity-main")
github_dataset <- read.csv("github_top_repositories.csv")

summary(github_dataset)
str(github_dataset)
```
## Data cleaning
```{r}
# check for na values
colSums(is.na(github_dataset))
sum(github_dataset$Description == "")
sum(github_dataset$Primary.Language == "None")
```
```{r}
library(dplyr) 
github_dataset_unique <- github_dataset %>% #remove duplicates
  distinct(Full.Name, .keep_all = TRUE)

nrow(github_dataset)        
nrow(github_dataset_unique)
```
# EDA
```{r}
library(ggplot2)
library(tidyr)
library(car)

#categorical
github_dataset_unique %>%
  filter(Primary.Language != "None" & Primary.Language != "") %>%
  count(Primary.Language, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(Primary.Language, n), y = n, fill = Primary.Language)) +
  geom_col(color = "white") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Top 10 Most Frequent Primary Languages",
    x = "Programming Language",
    y = "Repository Count"
  )

# 1. Select and rename continuous variables
continuous_data <- github_dataset_unique %>%
  select(
    stars  = contains("Stars"),
    forks  = contains("Forks"),
    issues = contains("Issues"),
    size   = contains("Size"),
    watchers = contains("Watchers.count")
  )

# 2. Inspect the data
summary(continuous_data)

# 3. Plot the scatter plot matrix
scatterplotMatrix(continuous_data)

```
You can see there is a correlation between forks vs stars but based on the distribution on the continuous variables, they are very right skewed. Watcher and stars also have the same values because "watching" a repository meant both bookmarking it (starring) and subscribing to its notifications. Later on, GitHub separated these into Starring (bookmarking/liking) and Watching (subscribing to updates). Basically the field watchers_count is an older, legacy field that GitHub kept for backwards compatibility, which actually just returns the exact same number as the stargazers/stars so we chose to drop watchers.

# dealing with the extreme right skew
```{r}
log_continuous_data <- continuous_data %>%
  mutate(
    stars  = log1p(stars),
    forks  = log1p(forks),
    issues = log1p(issues),
    size   = log1p(size)
  )

# 3. Inspect the newly transformed data
# You'll notice the mean and median are now much closer together (symmetrical!)
summary(log_continuous_data)

# 4. Plot the new, log-transformed scatter plot matrix
# The data points will no longer be squished in the bottom-left corner.
# Instead, they will be beautifully spread out, allowing you to see linear relationships!
scatterplotMatrix(log_continuous_data)
```
```{r}
regression_data <- github_dataset_unique %>%
  select(
    stars  = Stars.Count,
    forks  = Forks.Count,
    issues = Open.Issues.Count,
    size   = Size..KB.
  )

model_multiple <- lm(stars ~ ., data = regression_data)
summary(model_multiple)

regression_data$predicted <- predict(model_multiple)

ggplot(regression_data, aes(x = predicted, y = stars)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  # This orange line represents a perfect prediction
  geom_abline(slope = 1, intercept = 0, color = "darkorange", linewidth = 1.2) +
  theme_minimal() +
  labs(
    title = "Multiple Linear Regression: Actual vs. Predicted Stars",
    subtitle = "Predictors: Forks, Issues, and Size (Log-Log Scale)",
    x = "Predicted log(Stars + 1)",
    y = "Actual log(Stars + 1)"
  )
```
