---
title: "Culminating Activity"
author: Avryl Joie Arranguez, Dustin Jesse Balansag, Leira Bengil, John Andrew Chua,
  Jian Bryce Machacon
date: "2026-05-03"
output:
  html_document:
    df_print: paged
---
# Initializing & inspecting the Dataset
```{r}
setwd("C:/Users/USER/OneDrive/Desktop/web projects/R/github-R")
github_dataset <- read.csv("github_top_repositories.csv")

summary(github_dataset)
str(github_dataset)
```
## Data cleaning
```{r}
# check for na values
colSums(is.na(github_dataset))
sum(github_dataset$Description == "")
sum(github_dataset$Primary.Language == "None")
```
```{r}
library(dplyr) 
github_dataset_unique <- github_dataset %>% #remove duplicates
  distinct(Full.Name, .keep_all = TRUE)

nrow(github_dataset)        
nrow(github_dataset_unique)
```
# EDA
```{r}
library(ggplot2)
library(tidyr)
library(car)

#categorical
github_dataset_unique %>%
  filter(Primary.Language != "None" & Primary.Language != "") %>%
  count(Primary.Language, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(Primary.Language, n), y = n, fill = Primary.Language)) +
  geom_col(color = "white") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Top 10 Most Frequent Primary Languages",
    x = "Programming Language",
    y = "Repository Count"
  )

# 1. Select and rename continuous variables
continuous_data <- github_dataset_unique %>%
  select(
    stars  = contains("Stars"),
    forks  = contains("Forks"),
    issues = contains("Issues"),
    size   = contains("Size"),
    watchers = contains("Watchers.count")
  )

# 2. Inspect the data
summary(continuous_data)

# 3. Plot the scatter plot matrix
scatterplotMatrix(continuous_data)

```
You can see there is a correlation between forks vs stars but based on the distribution on the continuous variables, they are very right skewed. Watcher and stars also have the same values because "watching" a repository meant both bookmarking it (starring) and subscribing to its notifications. Later on, GitHub separated these into Starring (bookmarking/liking) and Watching (subscribing to updates). Basically the field watchers_count is an older, legacy field that GitHub kept for backwards compatibility, which actually just returns the exact same number as the stargazers/stars so we chose to drop watchers.

# dealing with the extreme right skew
```{r}
log_continuous_data <- continuous_data %>%
  mutate(
    stars  = log1p(stars),
    forks  = log1p(forks),
    issues = log1p(issues),
    size   = log1p(size)
  )

# 3. Inspect the newly transformed data
# You'll notice the mean and median are now much closer together (symmetrical!)
summary(log_continuous_data)

# 4. Plot the new, log-transformed scatter plot matrix
# The data points will no longer be squished in the bottom-left corner.
# Instead, they will be beautifully spread out, allowing you to see linear relationships!
scatterplotMatrix(log_continuous_data)
```
```{r}
regression_data <- github_dataset_unique %>%
  select(
    stars  = Stars.Count,
    forks  = Forks.Count,
    issues = Open.Issues.Count,
    size   = Size..KB.
  )

model_multiple <- lm(stars ~ ., data = regression_data)
summary(model_multiple)

regression_data$predicted <- predict(model_multiple)

ggplot(regression_data, aes(x = predicted, y = stars)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  # This orange line represents a perfect prediction
  geom_abline(slope = 1, intercept = 0, color = "darkorange", linewidth = 1.2) +
  theme_minimal() +
  labs(
    title = "Multiple Linear Regression: Actual vs. Predicted Stars",
    subtitle = "Predictors: Forks, Issues, and Size (Log-Log Scale)",
    x = "Predicted log(Stars + 1)",
    y = "Actual log(Stars + 1)"
  )
```

```{r}
library(ggplot2)
library(dplyr)

# 1. Prepare the data (using unique dataset already created)
regression_data <- github_dataset_unique %>%
  select(
    stars  = Stars.Count,
    forks  = Forks.Count,
    issues = Open.Issues.Count,
    size   = Size..KB.
  )

# 2. Split the data 80/20
set.seed(42) # For reproducibility
sample_size <- floor(0.80 * nrow(regression_data))
train_indices <- sample(seq_len(nrow(regression_data)), size = sample_size)

train_data <- regression_data[train_indices, ]
test_data  <- regression_data[-train_indices, ]

# 3. Train the Multiple Linear Regression model on training data
model_raw <- lm(stars ~ forks + issues + size, data = train_data)

# Show model coefficients and training performance
summary(model_raw)

# 4. Predict on the test set (the 20% holdout)
test_data$predicted_stars <- predict(model_raw, newdata = test_data)

# 5. Evaluate performance on the test set
# Calculate Mean Absolute Error (MAE)
mae <- mean(abs(test_data$predicted_stars - test_data$stars))
# Calculate R-squared for test set
rss <- sum((test_data$predicted_stars - test_data$stars)^2)
tss <- sum((test_data$stars - mean(test_data$stars))^2)
test_r2 <- 1 - (rss/tss)

print(paste("Mean Absolute Error (MAE):", round(mae, 2)))
print(paste("Test Set R-squared:", round(test_r2, 4)))

# 6. Visualization: Actual vs. Predicted
ggplot(test_data, aes(x = predicted_stars, y = stars)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  # Red dashed line represents a perfect 1:1 prediction
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Predictive Analytics: Actual vs. Predicted Stars",
    subtitle = "Tested on 20% Holdout Data (Raw Scale)",
    x = "Predicted Stars",
    y = "Actual Stars"
  )
```
## Analysis of Predictive Results
Following the 80/20 train-test split requested, the Multiple Linear Regression model was trained on 80% of the data and validated against the remaining 20% holdout set. Despite the extreme right skew of the raw data, the model yielded strong predictive insights.

###1. Model Equation and Coefficient Interpretation
The resulting regression equation is:
$$\text{Stars} = 8,365 + (3.758 \times \text{Forks}) + (1.158 \times \text{Issues}) - (0.0015 \times \text{Size})$$
Forks (Estimate: 3.758): This is the most influential predictor. For every single additional fork, the model predicts an increase of approximately 3.76 stars. This high coefficient suggests that developers who fork a repository (indicating intent to contribute or build upon the code) are highly likely to have also "starred" the project.

Open Issues (Estimate: 1.158): Interestingly, a higher count of open issues is associated with a higher star count. While issues can represent bugs, in top-tier repositories, they often signify a high level of user engagement and an active community.

Size (Estimate: -0.0015): The size of the repository has a negligible and slightly negative impact. This suggests that the physical scale of the codebase is not a driver of popularity; rather, community interaction (forks/issues) is what determines star counts.

## 2. Statistical Significance

All predictors yielded $p$-values significantly lower than the standard alpha of 0.05 (specifically $p < 0.001$ for Forks and Size, and $p < 0.01$ for Issues). This confirms that the relationship between these independent variables and the star count is statistically significant and not due to random chance.

## 3. Predictive Accuracy and Generalization
Test Set R-squared (0.6455): The model achieved an R-squared of 64.55% on the test set. Remarkably, this is higher than the training R-squared (56.33%), indicating that the model generalizes exceptionally well and is not overfitted to the training data. It effectively explains nearly two-thirds of the variance in stars.

Mean Absolute Error (MAE: 8,921.31): On average, our predictions deviate from the actual star count by approximately 8,921 stars. Given that the dataset includes "superstar" repositories with counts exceeding 100,000, this level of error is acceptable for a raw-scale linear model.
## 4. Conclusion for Predictive Analytics
The visualization of Actual vs. Predicted Stars shows a strong linear trend along the identity line (the red dashed line). While the raw values lead to some large residuals for extreme outliers (viral repositories), the model successfully captures the underlying trend of the GitHub ecosystem: Engagement (Forks) and Activity (Issues) are the primary engines of repository popularity.
